Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-26 arxiv/examples.tex
start sec 3/6

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
%We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
%The results in Section \ref{sec_general} show when multi-task learning outperforms single-task learning for general task covariance matrices and ground truth model parameters.
%In particular, the general conditions depend on specific properties of task data such as the covariate shift matrix.
%However, the conditions given by the limiting bias and variance of multi-task learning can be difficult to interpret.
%The goal of this section is to address this issue.
%We interpret the limiting bias and variance equations in an isotropic covariance setting that is described below and a simplified covariate-shifted setting.
%We show that three properties of task data affect the performance of multi-task learning as follows.
%(i) \textit{Task similarity}: We explain the phenomenon of negative transfer precisely as tasks become more different.
%(ii) \textit{Sample ratio}: We further explain a curious phenomenon where increasing the sample ratio between the source and target task helps initially, but hurts eventually.
%(iii) \textit{Covariate shift}: Finally, we show that as the sample size of the source task increases, we show that the covariate shift worsens the performance of the multi-task estimator.
%Finally, we provide a extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.
We illustrate our main results, Theorem \ref{thm_many_tasks}, Theorem \ref{thm main RMT} and Corollary \ref{thm_main_informal}, in a simplified setting.
For this setting, we simulate several empirical phenomena that are well-known in hard parameter sharing approaches.
We use our results to explain the empirical phenomena.

>>>>

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
%We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
%The results in Section \ref{sec_general} show when multi-task learning outperforms single-task learning for general task covariance matrices and ground truth model parameters.
%In particular, the general conditions depend on specific properties of task data such as the covariate shift matrix.
%However, the conditions given by the limiting bias and variance of multi-task learning can be difficult to interpret.
%The goal of this section is to address this issue.
%We interpret the limiting bias and variance equations in an isotropic covariance setting that is described below and a simplified covariate-shifted setting.
%We show that three properties of task data affect the performance of multi-task learning as follows.
%(i) \textit{Task similarity}: We explain the phenomenon of negative transfer precisely as tasks become more different.
%(ii) \textit{Sample ratio}: We further explain a curious phenomenon where increasing the sample ratio between the source and target task helps initially, but hurts eventually.
%(iii) \textit{Covariate shift}: Finally, we show that as the sample size of the source task increases, we show that the covariate shift worsens the performance of the multi-task estimator.
%Finally, we provide a extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.
We illustrate our main results in simplified settings.
We explain several empirical phenomena that have often been observed when hard parameter sharing approaches are used.

### 2020-09-26 arxiv/experiments.tex
wip
edits
start sec 3/6

\section{Experiments}

@@

\begin{table}[!b]
\begin{minipage}[t]{.58\textwidth}
	\vspace{-0.1in}
	\centering
  \begin{tabular}{c c c c c}
	\toprule
		\multirow{2}{*}{{\bf Threshold}}  & \multicolumn{2}{c}{{\bf Sentiment
		analysis}} & \multicolumn{2}{c}{{\bf ChestX-ray14}} \\
		& Precision &  Recall & Precision &  Recall \\
		\cmidrule(lr){1-1} \cmidrule(lr){2-3} \cmidrule(lr){4-5}
		0.0 & 0.596 & 1.000 & 0.593 & 1.000 \\
		0.1 & \textbf{0.756} & \textbf{0.388} & \textbf{0.738} & \textbf{0.462} \\
		0.2 & 0.919 & 0.065 & 0.875 & 0.044 \\
		% 0.3 & 1.000 & 0.004 &     - &     - \\
	\bottomrule
	\end{tabular}
	\vspace{0.1in}
	\captionof{table}{Single-task learning results can help predict postive or negative transfer in multi-task learning.}
	\label{tab:mtl_better_than_stl}
\end{minipage}%
\quad
	\begin{minipage}[t]{.40\textwidth}
	\vspace{-0.1in}
	\centering
	\begin{tabular}{c c c}
		\toprule
		% \multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\begin{minipage}{1.2in}\begin{center}
		% Sentiment\\ analysis\end{center}\end{minipage}} \\
		\multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\bf Sentiment analysis} \\
		% \cmidrule(lr){2-3}
		& all tasks & w/o TREC \\
		\midrule
		{\bf MLP}  & 31\% & 29\% \\
		{\bf LSTM} & 35\% & 34\% \\
		{\bf CNN}  & 30\% & 28\% \\
		\bottomrule
		\end{tabular}
	\vspace{0.1in}
	\captionof{table}{Efficiency of incremental training compared to baseline MTL.}
	\label{tab:taskonomy}
\end{minipage}
\end{table}

@@

%\textbf{The training procedure and baseline.}
For all models, we use a shared module for all tasks and assign a separate output layer on top of the shared module for each task.
The baseline training schedule for MTL is the round-robin training schedule.
%We compare our incremental training schedule of Algorithm \ref{alg_inc_train} to the round-robin training schedule.
We measure the test accuracy of predicting a target task.
We measure computational cost by summing over all epochs the number of samples used in every epoch.

\subsection{Simulation Studies}

\subsection{Validating the Theoretical Implication}\label{sec_exp_ab}

We first validate our theoretical results in Section \ref{sec_similarity} and \ref{sec_data_size}.
In Figure \ref{fig_ab_sim}, we compare the performance training with a semantically similar task versus a dissimilar task with a target task.
We select each task pair based on our domain knowledge.
We observe that adding a similar task helps the target task whereas adding a dissimilar task hurts.
In Figure \ref{fig_ab_data}, we validate that adding more source samples does not always improve performance on the target task.
Finally, we validate the algorithmic consequence of Section \ref{sec_covshift}.
In Figure \ref{fig_ab_cov}, we measure the performance gains from performing the alignment procedure proposed in \cite{WZR20} minus baseline MTL.
We average the results over all 15 task pairs.
The result shows that as the source samples increases, the alignment procedure shows a bigger improvement over MTL.
The rest of experimental procedures are left to Appendix \ref{app_it}.

\subsection{Validating the Algorithmic Consequences}\label{sec_exp_res}

\textit{Predicting transfer effect via STL results.}
We show that the single-task based metric proposed in Section \ref{sec_similarity} can predict positive or negative transfer in MTL.
A common challenge in the study of MTL is that the results can be hard to understand.
It is difficult to predict when MTL performs well without running extensive trials.
Our insight is that we can use STL results to help understand MTL results.
Table \ref{tab:mtl_better_than_stl} shows the result on both the sentiment analysis and the ChestX-ray14 tasks.
We find that using a threshold of $\tau = 0.1$, the STL results correctly predict positive or negative transfer with $75.6\%$ precision and $38.8\%$ recall among $30$ times $5$ (random seeds) task pairs!
We observe similar results for $91$ task pairs from the ChestX-ray14 dataset.
%The results show that STL results are indicative of MTL results.

\textit{Mitigating negative transfer via incremental training.}
First, we show that our proposed incremental training schedule (Algorithm \ref{alg_inc_train}) can help mitigate negative transfer for predicting a particular target task.
Over all $15$ pairs from the sentiment analysis tasks, we find that Algorithm \ref{alg_inc_train} requires only $45\%$ of the computational cost to achieve similar performance on the target task, compared to the MTL baseline.
Our insight is that since adding more samples from the source task does not always help, we can improve efficiency by adding source samples \textit{incrementally} during training.

@@

\subsection{Missing Details from the Experiments}\label{app_experiments}

\subsection{Synthetic Settings}\label{app_synthetic}
	We describe the parameter settings used in the synthetic experiments (Figure \ref{fig_model_shift_phasetrans}).
	For all three synthetic experiments, we set the input feature dimension as $p = 200$.
	\squishlist
		\item Task similarity: We use $\rho_1 = 90, \rho_2 = 30$ for sample sizes.
	The noise level is $\sigma = 5$.
	We set $\kappa = 1$ and vary model distance $d$ from $0.01$ to $0.2$.
	The $x$-axis measures the function $\Phi(\beta_1, \beta_2)$ described in Section \ref{sec_similarity}.
	The $y$-axis measures the improvement of prediction loss using MTL compared to STL.
		\item Sample size: We set $\kappa = 1$ and $d = 0.02$ for model distance.
	We use $\rho_2 = 500$ and vary $\rho_1$ from $50$ to $800$ for sample sizes.
		\item Covariate shift: We set $\kappa = 1$ and $d = 0$.
	We set $\rho_2 = 4$ and vary $\rho_1$ from $5$ to $25$ for sample sizes.
	We use the scale parameter $\lambda = 1$ for the curve without covariate shift and $\lambda = 2$ for the curve with covariate shift (cf. Section \ref{sec_covshift}).
	Without loss of generality, we have rescaled the $x$ and $y$ axis for the ease of presentation.
	This can be achieved similarly by rescaling the task data.
	\squishend

\subsection{Image and Text Classification Settings}\label{app_it}

For the case of six tasks, we extend Algorithm \ref{alg_inc_train} accordingly.
Initially, we use $5\%$ of samples from SST and $50\%$ of samples from the other tasks.
We add $19\%$ of samples from SST and $5\%$ of samples from the other tasks for every $2$ epochs.
We run $20$ epochs in total.

@@

>>>>

\section{Numerical Results}

@@

\begin{table}[!b]
%\begin{minipage}[t]{.58\textwidth}
%	\vspace{-0.1in}
%	\centering
%  \begin{tabular}{c c c c c}
%	\toprule
%		\multirow{2}{*}{{\bf Threshold}}  & \multicolumn{2}{c}{{\bf Sentiment
%		analysis}} & \multicolumn{2}{c}{{\bf ChestX-ray14}} \\
%		& Precision &  Recall & Precision &  Recall \\
%		\cmidrule(lr){1-1} \cmidrule(lr){2-3} \cmidrule(lr){4-5}
%		0.0 & 0.596 & 1.000 & 0.593 & 1.000 \\
%		0.1 & \textbf{0.756} & \textbf{0.388} & \textbf{0.738} & \textbf{0.462} \\
%		0.2 & 0.919 & 0.065 & 0.875 & 0.044 \\
		% 0.3 & 1.000 & 0.004 &     - &     - \\
%	\bottomrule
%	\end{tabular}
%	\vspace{0.1in}
%	\captionof{table}{Single-task learning results can help predict postive or negative transfer in multi-task learning.}
%	\label{tab:mtl_better_than_stl}
%\end{minipage}%
%\quad
	\begin{minipage}[t]{.40\textwidth}
	\vspace{-0.1in}
	\centering
	\begin{tabular}{c c c}
		\toprule
		% \multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\begin{minipage}{1.2in}\begin{center}
		% Sentiment\\ analysis\end{center}\end{minipage}} \\
		\multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\bf Sentiment analysis} \\
		% \cmidrule(lr){2-3}
		& all tasks & w/o TREC \\
		\midrule
		{\bf MLP}  & 31\% & 29\% \\
		{\bf LSTM} & 35\% & 34\% \\
		{\bf CNN}  & 30\% & 28\% \\
		\bottomrule
		\end{tabular}
	\vspace{0.1in}
	\captionof{table}{Efficiency of incremental training compared to baseline MTL.}
	\label{tab:taskonomy}
\end{minipage}
\end{table}

@@

\paragraph{Synthetic Settings}\label{app_synthetic}
	We describe the parameter settings used in the synthetic experiments (Figure \ref{fig_model_shift_phasetrans}).
	For all three synthetic experiments, we set the input feature dimension as $p = 200$.
	\squishlist
		\item Task similarity: We use $\rho_1 = 90, \rho_2 = 30$ for sample sizes.
	The noise level is $\sigma = 5$.
	We set $\kappa = 1$ and vary model distance $d$ from $0.01$ to $0.2$.
	The $x$-axis measures the function $\Phi(\beta_1, \beta_2)$ described in Section \ref{sec_similarity}.
	The $y$-axis measures the improvement of prediction loss using MTL compared to STL.
		\item Sample size: We set $\kappa = 1$ and $d = 0.02$ for model distance.
	We use $\rho_2 = 500$ and vary $\rho_1$ from $50$ to $800$ for sample sizes.
		\item Covariate shift: We set $\kappa = 1$ and $d = 0$.
	We set $\rho_2 = 4$ and vary $\rho_1$ from $5$ to $25$ for sample sizes.
	We use the scale parameter $\lambda = 1$ for the curve without covariate shift and $\lambda = 2$ for the curve with covariate shift (cf. Section \ref{sec_covshift}).
	Without loss of generality, we have rescaled the $x$ and $y$ axis for the ease of presentation.
	This can be achieved similarly by rescaling the task data.
	\squishend

\paragraph{Image and Text Classification Settings}\label{app_it}

For the case of six tasks, we extend Algorithm \ref{alg_inc_train} accordingly.
Initially, we use $5\%$ of samples from SST and $50\%$ of samples from the other tasks.
We add $19\%$ of samples from SST and $5\%$ of samples from the other tasks for every $2$ epochs.
We run $20$ epochs in total.
%\textbf{The training procedure and baseline.}
For all models, we use a shared module for all tasks and assign a separate output layer on top of the shared module for each task.
The baseline training schedule for MTL is the round-robin training schedule.
%We compare our incremental training schedule of Algorithm \ref{alg_inc_train} to the round-robin training schedule.
We measure the test accuracy of predicting a target task.
We measure computational cost by summing over all epochs the number of samples used in every epoch.

%\subsection{Validating the Theoretical Results}\label{sec_exp_ab}
%We first validate our theoretical results in Section \ref{sec_similarity} and \ref{sec_data_size}.
%In Figure \ref{fig_ab_sim}, we compare the performance training with a semantically similar task versus a dissimilar task with a target task.
%We select each task pair based on our domain knowledge.
%We observe that adding a similar task helps the target task whereas adding a dissimilar task hurts.
%The rest of experimental procedures are left to Appendix \ref{app_it}.

\subsection{Sample Imbalance}\label{sec_exp_sample}

%\textit{Predicting transfer effect via STL results.}
%We show that the single-task based metric proposed in Section \ref{sec_similarity} can predict positive or negative transfer in MTL.
%A common challenge in the study of MTL is that the results can be hard to understand.
%It is difficult to predict when MTL performs well without running extensive trials.
%Our insight is that we can use STL results to help understand MTL results.
%Table \ref{tab:mtl_better_than_stl} shows the result on both the sentiment analysis and the ChestX-ray14 tasks.
%We find that using a threshold of $\tau = 0.1$, the STL results correctly predict positive or negative transfer with $75.6\%$ precision and $38.8\%$ recall among $30$ times $5$ (random seeds) task pairs!
%We observe similar results for $91$ task pairs from the ChestX-ray14 dataset.
%The results show that STL results are indicative of MTL results.

In Figure \ref{fig_ab_data}, we validate that adding more source samples does not always improve performance on the target task.
Finally, we validate the algorithmic consequence of Section \ref{sec_covshift}.

\textit{Mitigating negative transfer via incremental training.}
\noindent\textit{Mitigating negative transfer.}
An interesting consequence of Proposition \ref{prop_data_size} is that $L(\hat{\beta}_t^{\MTL})$ is not monotone in $\rho_1$.
In particular, Figure \ref{fig_size} (and our analysis) shows that $L(\hat{\beta}_t^{\MTL})$ behaves as a quadratic function over $\rho_1$.
More generally, depending on how large $\Psi(\beta_1, \beta_2)$ is, $L(\hat{\beta}_t^{\MTL})$ may also be monotonically increasing or decreasing.
Based on this insight, we propose an incremental optimization schedule to improve MTL training efficiency.
\squishlist
	\item We divide the source task data into $S$ batches.
	For $S$ rounds, we incrementally add the source task data by adding one batch at a time.
	\item After training $T$ epochs, if the validation accuracy becomes worse than the previous round's result, we terminate.
	Algorithm \ref{alg_inc_train} in Appendix \ref{app_experiments} describes the procedure in detail.
\squishend

First, we show that our proposed incremental training schedule (Algorithm \ref{alg_inc_train}) can help mitigate negative transfer for predicting a particular target task.
Over all $15$ pairs from the sentiment analysis tasks, we find that Algorithm \ref{alg_inc_train} requires only $45\%$ of the computational cost to achieve similar performance on the target task, compared to the MTL baseline.
Our insight is that since adding more samples from the source task does not always help, we can improve efficiency by adding source samples \textit{incrementally} during training.

@@

\subsection{Covariate Shift}

\noindent\textit{Covariance alignemnt.}
Our observation highlights the need to correct covariate shift when $n_1 / n_2$ is large.
Hence for such settings, we expect procedures that aim at correcting covariate shift to provide more significant gains.
We consider a covariance alignment procedure proposed in \cite{WZR20}, which is designed for the purpose of correcting covariate shift.
The idea is to add an alignment module between the input and the shared module $B$.
This new module is then trained together with $B$ and the output layers.
We validate our insight on this procedure in the experiments.

In Figure \ref{fig_ab_cov}, we measure the performance gains from performing the alignment procedure proposed in \cite{WZR20} minus baseline MTL.
We average the results over all 15 task pairs.
The result shows that as the source samples increases, the alignment procedure shows a bigger improvement over MTL.

@@

\subsection{Sample Efficiency}

\begin{table}
	\begin{center}
		\begin{tabular}{c c c c c}
			\toprule
			\multirow{2}{*}{{\bf Models}} & \multicolumn{2}{c}{\begin{minipage}{1.1in}\begin{center}
				                                                                          MR, SST, SUBJ, CR, MPQA, TREC\end{center}\end{minipage}} & \multicolumn{2}{c}{\begin{minipage}{1.1in}\begin{center}MR, SST, SUBJ, CR, MPQA\end{center}\end{minipage}} \\
			\cmidrule(lr){2-3} \cmidrule(lr){4-5}
			& {\bf Stanford} & {\bf Alignment} & {\bf Stanford} & {\bf Alignment} \\
			\midrule
			{\bf MLP}  & > 100\% & 39\% & 25\% & 25\% \\
			{\bf LSTM} & 36\% & 36\% & 28\% & 25\% \\
			% {\bf CNN}  & 76\% & - & 32\% & -\\
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Taskonomy experiment.}
	\label{tab:taskonomy}
\end{table}
